A lottery model of density-dependent selection

### 2017-11-08 manuscript.tex
l

Fitness is typically represented in heavily simplified terms in evolutionary genetics, often using constant selection coefficients

@@

MacArthur's $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is famous for its early role in life-history theory \citep{charlesworth_1994,reznick_2002}, and for initiating a sizeable literature on density-dependent selection \citep{christiansen_2004}, which contains some of earliest treatments of the interplay between ecology and evolution \citep{roughgarden_1979,travis_2013}. The central claim of the $r$/$K$ scheme is that in ``crowded populations ... fitness is $K$'' ($K$ will be defined below) \citep[pp. 149]{macarthur_1967}. Remarkably, not only does MacArthur's argument for this claim seem to apply to essentially any population subject to density-regulated growth, the claim is also incompatible with the standard representation of selection in evolutionary biology. Since the general version of the ``fitness is $K$'' claim has been somewhat obscured by the widespread use of single-locus variants of the logistic model in the density-dependent selection literature, as well as well as the problems of the $r$/$K$ life-history literature \cite[pp. 265]{charlesworth_1994}, we briefly review MacArthur's argument before proceeding.

MacArthur considers two types subject to density-dependent population growth described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2), \label{eq:macgeneral}
\end{equation}
where $f_1$ and $f_2$ are strictly decreasing functions of the population densities $n_1$ and $n_2$. Plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$, it becomes clear that one type will be excluded unless it is able to grow at greater densities than the other type in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}). The term ``$K$-selection'' comes from four ``carrying capacities'' defined by $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$. These $K$ values summarize the nullcline structure provided that the nullclines do not curve too much. Note that only $K_{11}$ and $K_{22}$ are saturation densities analogous to ``$K$'' in the logistic model; $K_{12}$ and $K_{21}$ represent the effects of competition between types. For instance, under Lotka-Volterra competition we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$, where $\alpha_{11}=1/K_{11}$ measures intra-type competitive effects, and $\alpha_{12}=1/K_{12}$ measures inter-type competitive effects. In the case of exploitation competition for consumable resources, intra- and inter-type competition are connected via resource use ``efficiency'', but obviously this need not be true more generally \citep{gill_1974,case_1974}. Nevertheless, bearing in mind MacArthur's broad use of ``$K$'', it seems hard to disagree that ``fitness is $K$'' i.e. that selection in crowded populations either favors the ability to keep growing at ever higher densities, or the ability to stop the growth of competitors at lower densities.

\begin{figure}
\centering
\includegraphics[scale=0.8]{Kplot.pdf}
\caption{\label{fig:Ksel} (a) MacArthur's dynamical argument for why ``fitness is $K$'' in crowded environments. The nullclines where $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ need not be linear, but the ``carrying capacities'' $K$ are then less informative about the nullcline structure. Assuming these nullclines exist, exclusion is certain unless a type's nullcline extends out to greater densities in some region (here the outcome depends on initial condition).}
\end{figure}

By contrast, in the standard ``relative fitness'' representation of selection, each type is assigned a selection coefficient describing the change in its relative proportion over time (e.g. \cite[pp. 468]{barton_2007}), with no consideration of population density. While this can describe selection in an uncrowded, expanding population, most commonly (e.g. in the Wright-Fisher model) it is assumed that total population density $N$ is fixed. In the latter case, the population is crowded in the sense that one type must displace another to grow. Selection happens along a line defined by $n_1+n_2=N$ and population density is otherwise irrelevant. Given this inconsistency with the $r$/$K$ scheme, the constant-$N$, shifting-frequency picture of selection does not seem compatible with the extremely broad variety of absolute population growth dynamics represented by Eq.~\eqref{eq:macgeneral}.

@@

>>>>

@@

A variety of different measures of fitness are used to quantify the effects of selection. Some widely used examples in evolutionary ecology are expected lifetime reproductive ratio $R_0$,  intrinsic growth rate $r$, saturation population density (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where attention is generally restricted to relative genotypic proportions \cite[pp. 468]{barton_2007}.

In principle these fitness measures should all be grounded in population ecology, because ultimately selection operates through the environmentally-dependent differential reproduction and survival of individuals \citep{metcalf_2007}. Widely-used ``absolute'' measures of fitness like $r$ and $K$ are derived directly from population ecology. By contrast, relative fitness is almost entirely divorced from population ecology. This discrepancy runs deeper than a simple difference in emphasis on absolute densities rather than relative frequencies.

In uncrowded populations there is no discrepancy. Relative fitness simply represents  differences in $r$ ($r$-selection) \citep[pp. 26]{crow_1970}. The discrepancy emerges in crowded populations. Then the relative fitness description of selection assumes that $N$ is fixed, and fitness strictly involves the ability to increase in frequency at the expense of others (e.g. in the Wright-Fisher model). On the other hand, from quite general population ecological considerations, MacArthur famously argued that ``fitness is $K$'' in crowded populations ($K$-selection) \citep[pp. 149]{macarthur_1967}. To appreciate the import of MacArthur's argument, and see the nature of the discrepancy between these views of selection, we now briefly review MacArthur's argument.

MacArthur considers two types subject to density-dependent population growth described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2), \label{eq:macgeneral}
\end{equation}
where $f_1$ and $f_2$ are strictly decreasing functions of the population densities $n_1$ and $n_2$. Plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$, it can be seen that one type will be excluded unless it is able to grow at higher densities than the other type in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a). MacArthur defines four ``carrying capacities'' generally labeled $K$, as the intersection points of the nullclines with the $n_1$ and $n_2$ axes (i.e. $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$). Assuming that the nullclines are linear, these $K$ values determine whether a region of higher-density growth exists for each type.

Note that only $K_{11}$ and $K_{22}$ are saturation densities analogous to ``$K$'' in the logistic model; $K_{12}$ and $K_{21}$ represent the effects of competition between types. For instance, under Lotka-Volterra competition we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$. Then the nullclines are linear and $\alpha_{11}=1/K_{11}$ measures intra-type competitive effects, while $\alpha_{12}=1/K_{12}$ measures inter-type competitive effects.  Thus, for MacArthur ``fitness is $K$'' in the sense that selection in crowded populations either favors the ability to keep growing at ever higher densities, or the ability to stop the growth of competitors at lower densities. This general idea applies even if the nullclines are so curved that the ``$K$'' values themselves do not give much information about the regions of high-density growth.

\begin{figure}
\centering
\includegraphics[scale=0.8]{Kplot.pdf}
\caption{\label{fig:Ksel} (a) MacArthur's dynamical argument for why ``fitness is $K$'' in crowded environments, using a Lotka-Volterra form for Eq. \eqref{eq:macgeneral}. Assuming the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ exist, exclusion is certain unless a type's nullcline extends out to greater densities in some region. In this example, the outcome depends on initial condition. (b) The constant-$N$, relative fitness description of selection.}
\end{figure}

Clearly MacArthur's claim is not compatible with the constant-$N$, relative fitness description of selection, in which $n_1$ and $n_2$ change along a line defined by $n_1+n_2=N$, and population density is otherwise irrelevant (Fig.~\ref{fig:Ksel}b). In this case, all four $K$ values are equal to $N$ [not!], and the $f_1$ and $f_2$ nullclines do not even exist. Since $N$ is fixed, it is not possible to persist at higher densities than $N$, nor is it possible to suppress the density of a competitor at densities lower than $N$

Given the extremely broad variety of absolute population growth dynamics represented by Eq.~\eqref{eq:macgeneral}, at this point we might start to doubt that the relative fitness, constant=$N$ description of selection has a plausible basis in population ecology. In particular, the uncoupling of selection and population demography could be problematic, because they are both driven by the same birth/death events.

@@

%In the case of exploitation competition for consumable resources, intra- and inter-type competition are connected via resource use ``efficiency'', but obviously this need not be true more generally \citep{gill_1974,case_1974}.
